Inference for Trans-dimensional Bayesian Models with Diffusive Nested Sampling

### 2014-08-28 Documents/Paper/ms.tex
Working on galaxyfield section

\section{Sinusoidal example}
In this section we demonstrate a seemingly simple example which exhibits
a phase transition, making a standard MCMC approach difficult. Suppose
a signal is composed of $N$ sinusoids, of different periods, amplitudes,
and phases. The signal is observed at various times $\{t_i\}_{i=1}^N$ with
noise, and we want to use the resulting data to infer the number of sinusoids
$N$, along with the periods $\{T_i\}_{i=1}^N$, amplitudes $\{A_i\}_{i=1}^N$,
and phases $\{\phi_i\}_{i=1}^N$.
This kind of model has many applications, and has been solved analytically
under certain sets of assumptions \citep[see e.g.][]{bretthorst}. Similar
models have been used in many different fields
\citep[e.g.][]{2005PhRvD..72b2001U, 2007ApJ...654..551B, 2003AIPC..659....3B}.
Note that this problem is also very closely related to the
problem of detecting exoplanet signals in radial velocity data, which has
attracted a lot of research attention in recent years
\citep[e.g.][]{2007ApJ...654..551B, gregory, fengji, 2011MNRAS.415.3462F}.

@@

%\section{``Transit'' Example}
%\section{``Asteroseismology'' Example}
\section{``Galaxy Field'' Example}
Source detection is an important problem in astrophysics. Given some data,
usually one or more noisy, blurred, images of the sky, we would like to know
how many objects $N$ (such as stars or galaxies) are in the image, and their
properties (such as flux, size, orientation). Various ad-hoc approaches exist,
covering a wide spectrum between ad-hoc approaches and principled inference
approaches
\citep[e.g.][]{irwin, sextractor, dolphot, 2003MNRAS.338..765H, starfield}.

In this section we apply
the DNS approach to a toy version of the problem of detecting and quantifying
extended objects such as galaxies in a noisy image.
We model each galaxy as a mixture of two concentric elliptical components, with
the following eight parameters. Firstly, two
parameters $x_c$ and $y_c$ describe the central position of the galaxy within
the image. The flux $f$ describes the total integral of the galaxy's
intensity profile. The axis ratio $q$ describes the ellipticity of the galaxy
and $\theta$ its orientation angle with respect to horizontal. The radius of the
bigger gaussian is a parameter $w$.
Finally
we include a ``radius ratio'' $u$ describing the ratio of the radius of the
smaller gaussian with respect to that of the bigger gaussian, and a parameter
$v$ describing the fraction of the total flux in the smaller gaussian
(therefore the fraction of the light in the bigger gaussian is $1-v$).
In a coordinate system aligned with the major axis of the ellipse, the
surface brightness profile of a ``galaxy'' is given by
\begin{eqnarray}
\rho(x, y) &=&
\frac{f(1-v)}{2\pi w^2}\exp\left[-\frac{1}{2w^2}\left(qx^2 + y^2/q\right)\right]\\
&&+\frac{fv}{2\pi (uw)^2}\exp\left[-\frac{1}{2(uw)^2}\left(qx^2 + y^2/q\right)\right].
\end{eqnarray}

For the interim prior for the total fluxes of the galaxies, and for the
widths of the galaxies, we use the Pareto
distribution. The probability density function for a variable $x$ given
hyperparameters $x_{\rm min}$ and $a$ is
\begin{eqnarray}
p(x | x_{\rm min}, a) &=&
\left\{
\begin{array}{lr}
\frac{a x_{\rm min}^{a}}{x^{a + 1}}, & x \geq x_{\rm min}\\
0, & \textnormal{otherwise.}
\end{array}
\right.
\end{eqnarray}
Since we are using this for both the fluxes and the widths of the galaxies,
there will be four hyperparameters in total for the fluxes and widths.

\begin{figure}
\begin{center}
\includegraphics[scale=0.7]{galaxyfield_data.pdf}
\caption{The simulated data for the GalaxyField example.
\label{fig:galaxyfield_data}}
\end{center}
\end{figure}

@@

In general, it is possible to compute the number of components that have been
affected by the proposal. If this is greater than or equal to $N$, we compute
the mock signal from scratch, otherwise we subtract the effect of those
components that have been removed and add in the effect of those that have been
added. In the C++ implementation of this work, available at
{\tt https://www.github.com/eggplantbren/RJObject/}, when a proposal takes
place, the difference between the old and new model is cached and is available
to the user, so that the likelihood can be updated rather than computed
from scratch.

>>>>

\section{Sinusoidal example}\label{sec:sinewaves}
In this section we demonstrate a seemingly simple example which exhibits
a phase transition, making a standard MCMC approach difficult. Suppose
a signal is composed of $N$ sinusoids, of different periods, amplitudes,
and phases. The signal is observed at various times $\{t_i\}_{i=1}^N$ with
noise, and we want to use the resulting data to infer the number of sinusoids
$N$, along with the periods $\{T_i\}_{i=1}^N$, amplitudes $\{A_i\}_{i=1}^N$,
and phases $\{\phi_i\}_{i=1}^N$.
This kind of model has many applications, and has been solved analytically
under certain sets of assumptions \citep[see e.g.][]{bretthorst}. Similar
models have been used in many different fields
\citep[e.g.][]{2005PhRvD..72b2001U, 2007ApJ...654..551B, 2003AIPC..659....3B}.
Note that this problem is also very closely related to the
problem of detecting exoplanet signals in radial velocity data, which has
attracted a lot of research attention in recent years
\citep[e.g.][]{2007ApJ...654..551B, gregory, fengji, 2011MNRAS.415.3462F}.

@@

%\section{``Transit'' Example}
%\section{``Asteroseismology'' Example}
\section{``Galaxy Field'' Example}
Source detection is an important problem in astrophysics. Given some data,
usually one or more noisy, blurred, images of the sky, we would like to know
how many objects $N$ (such as stars or galaxies) are in the image, and their
properties (such as flux, size, orientation). Various approaches exist,
covering a wide spectrum between ad-hoc approaches and principled inference
approaches
\citep[e.g.][]{irwin, sextractor, dolphot, 2003MNRAS.338..765H, starfield}.
The more principled techniques tend to be much more computationally
intensive, so should only be applied to interesting or
especially challenging subsets of astronomical imaging.

In this section we apply
the DNS approach to a toy version of the problem of detecting and quantifying
extended objects such as galaxies in a noisy image.
We model each galaxy as a mixture of two concentric elliptical gaussian
components, with
the following eight parameters. Firstly, two
parameters $x_c$ and $y_c$ describe the central position of the galaxy within
the image. The flux $f$ describes the total integral of the galaxy's
intensity profile. The axis ratio $q$ describes the ellipticity of the galaxy
and $\theta$ its orientation angle with respect to horizontal. The radius of the
bigger gaussian is a parameter $w$.
Finally
we include a ``radius ratio'' $u$ describing the ratio of the radius of the
smaller gaussian with respect to that of the bigger gaussian, and a parameter
$v$ describing the fraction of the total flux in the smaller gaussian
(therefore the fraction of the light in the bigger gaussian is $1-v$).
In a coordinate system aligned with the major axis of the ellipse, the
surface brightness profile of a ``galaxy'' is given by
\begin{eqnarray}
\rho(x, y) &=&
\frac{f(1-v)}{2\pi w^2}\exp\left[-\frac{1}{2w^2}\left(qx^2 + y^2/q\right)\right]\\
&&+\frac{fv}{2\pi (uw)^2}\exp\left[-\frac{1}{2(uw)^2}\left(qx^2 + y^2/q\right)\right].
\end{eqnarray}
We generated a simulated $200 \times 200$ pixel noisy image
(Figure~\ref{fig:galaxyfield_data}) containing 47 ``galaxies'',
in order to test the algorithm.

\begin{figure}
\begin{center}
\includegraphics[scale=0.7]{galaxyfield_data.pdf}
\caption{The simulated image for the GalaxyField example. The image is
$200\times 200$ pixels in size and contains $N=47$ ``galaxies''.
\label{fig:galaxyfield_data}}
\end{center}
\end{figure}

For the interim prior for the total fluxes of the galaxies, and for the
widths of the galaxies, we used the Pareto
distribution. The probability density function for a variable $x$ given
hyperparameters $x_{\rm min}$ and $a$ is
\begin{eqnarray}
p(x | x_{\rm min}, a) &=&
\left\{
\begin{array}{lr}
\frac{a x_{\rm min}^{a}}{x^{a + 1}}, & x \geq x_{\rm min}\\
0, & \textnormal{otherwise.}
\end{array}
\right.
\end{eqnarray}
Since we are using this for both the fluxes and the widths of the galaxies,
there will be four hyperparameters in total: two lower limits (one for the
fluxes and one for the widths) and two slopes. As in Section~\ref{sec:sinewaves}
we allowed the noise standard deviation $\sigma$ to be a free parameter with
a log-uniform prior between $10^{-3}$ and $10^3$.

@@

In general, it is possible to compute the number of components that have been
affected by the proposal. If this is greater than or equal to $N$, we compute
the mock signal from scratch, otherwise we subtract the effect of those
components that have been removed and add in the effect of those that have been
added. In the C++ implementation of this work, when a proposal takes
place, the difference between the old and new model is cached and is available
to the user, so that the likelihood can be updated rather than computed
from scratch.
